---
output:
  bookdown::gitbook:
  pdf_document: default
  html_document: default
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, results = 'hide', fig.show="hide",message=FALSE,warning=FALSE,cache=TRUE)
```



# Multiple covariates {#ch:multiple}

## Examples 

<b>Reading materials</b>:  Slides 101 - 110 in STA108_LinearRegression_S20.pdf. 

We will revisit the examples that have been studied in Chapter \@ref(ch:lmR).

### Advertising data 

We now consider all the covariates in the data set. The resulting model is
	\[
	y_i =  \beta_0 + \sum_{j=1}^3 x_{i,j} \beta_j  +  \epsilon_i, i=1,\ldots, 200,
	\]
	where $y_i$ is the `sales` (1 unit = 1000 dollars) for the $i$th entry, $x_{i,1}$ is the TV advertising budget for the $i$th entry,  $x_{i,2}$ is the radio advertising budget for the $i$th entry, $x_{i,3}$ is the newspaper advertising budget for the $i$th entry. In addition, we assume that the errors $\{\epsilon_i\}_{i=1}^{200}$ satisfy that $\epsilon_1,\ldots, \epsilon_200$ are independently and identically distributed (i.i.d.),  $\mathbb{E}[\epsilon_i]= 0$ for  $i=1,2,\ldots, 200$
and $\mathrm{var}(\epsilon_i)=\sigma^2$ for  $i=1,2,\ldots, 200$. Recall that we consider fixed design (i.e., $x_i$ is not random) in this course for simplicity. 

```{r}
dat.advertising=read.csv('./data/advertising.csv');
# Fit a multiple linear regression
fit.advertising = lm(sales~TV+radio+newspaper+1,data=dat.advertising); 
# Summarize the fitted results
summary(fit.advertising) 

fit.advertising.slr = lm(sales~TV+1,data=dat.advertising); 
# Summarize the fitted results
summary(fit.advertising.slr) 


```

How would you interpret the fitted results? How do they compare with the results from the simple linear regression in Chapter \@ref(lmR)?

### Flu shot

We will include `age` and `female`(gender) in the regression in addition to the treatment received. What can you conclude from the fitted results?


```{r}
dat.flu = read.table("./data/flu240.txt", header = TRUE)
# Fit a multiple linear regression
fit.flu= lm(outcome~treatment.received+age+female+1,data=dat.flu); 
# Summarize the fitted results
summary(fit.flu) 
```
In addition, we can fit a regression with treatment received as the outcome, where treatment assigned, age and gender are covariates.  What can you conclude now? 
```{r}
fit.trt= lm(treatment.received~treatment.assigned+age+female+1,data=dat.flu); 
# Summarize the fitted results
summary(fit.trt) 
```

### Project STAR


Project STAR is a stratified randomized experiment, where randomization happened within each school. Hence, we will add `schoolid2` as an additional factor in the regression. 

```{r}

library(AER)
data("STAR")
dat.STAR=STAR; # Just to be consistent
# Fit a simple linear regression
fit.STAR= lm(math2~as.factor(star2)+as.factor(schoolid2)+1,data=dat.STAR); 
# Summarize the fitted results
summary(fit.STAR)$coef[1:3,]

# We can no longer draw a fitted line here...

```




## Classification of variables 


<b>Reading materials</b>:  Slides 111 - 118 in STA108_LinearRegression_S20.pdf. 

We will use simulation to examine the roles of the variables. 

```{r}
source('sources.r')

simulate.one.instance.data<-function(){
n=100;
confounder = rnorm(n)*2;
precision = runif(n);
instrument = rbinom(n,size=1,prob=0.4);
trt=rnorm(n) + confounder*0.4-instrument;
y=trt*1-0.5*confounder+precision*0.7+0.5*rnorm(n);
out=data.frame(y=y,trt=trt,confounder=confounder, precision=precision, instrument=instrument)
return(out)
}

```

```{r}
## Effect if ignoring confounder:

simulate.one.instance<-function(){
  dat=simulate.one.instance.data();
  beta.hat=fit.linear.model(covariate=dat[,'trt'],outcome=dat$y)
  
  beta.hat.confounder=fit.linear.model(covariate=as.matrix(dat[,c('trt','confounder')]),outcome=dat$y)
  return(c(beta.hat[2],beta.hat.confounder[2]))
}
N.sim=10000;

set.seed(1)

sim.confounder=replicate(N.sim, simulate.one.instance());
apply(sim.confounder,1,mean) # The true value is 1
apply( (sim.confounder-1),1,function(x){ sum(x^2) }) # The mean squared error
```

```{r}
## Effect if ignoring precision variable:

simulate.one.instance<-function(){
  dat=simulate.one.instance.data();
  beta.hat=fit.linear.model(covariate=as.matrix(dat[,c('trt','confounder')]),outcome=dat$y)
  
  beta.hat.precision=fit.linear.model(covariate=as.matrix(dat[,c('trt','confounder','precision')]),outcome=dat$y)
  return(c(beta.hat[2],beta.hat.precision[2]))
}
N.sim=10000;

set.seed(1)

sim.precision=replicate(N.sim, simulate.one.instance());
apply(sim.precision,1,mean) # The true value is 1
apply( (sim.precision-1),1,function(x){ sum(x^2) }) # The mean squared error
```

```{r}
## The power of instruments
###  If the confounder is unobserved, we can use the two stage least squares method

simulate.one.instance<-function(){
  dat=simulate.one.instance.data();
  beta.hat.naive=fit.linear.model(covariate=as.matrix(dat[,'trt']),outcome=dat$y)
  
  yiv=fit.linear.model(covariate=as.matrix(dat[,c('instrument')]),outcome=dat$y)
  
  trtiv=fit.linear.model(covariate=as.matrix(dat[,c('instrument')]),outcome=dat$trt)
  
  
  return(c(beta.hat.naive[2],yiv[2]/trtiv[2]))
}
N.sim=10000;

set.seed(1)

sim.iv=replicate(N.sim, simulate.one.instance());
apply(sim.iv,1,mean) # The true value is 1
apply( (sim.iv-1),1,function(x){ sum(x^2) }) # The mean squared error
```


## Least squares estimation 

<b>Reading materials</b>:  Slides 119 - 141 in STA108_LinearRegression_S20.pdf. 



```{r}
# Consider another data set with a bit more covariates 
library(AER)
data("Fatalities")
dat.fatalities = Fatalities
# This dataset contains 34 variables
# It is actually a longitudinal/panel data
# We will ignore the spatial temporal structure in this class

y=dat.fatalities$fatal; #Number of vehicle fatalities.
X=as.matrix(dat.fatalities[,c('spirits','unemp','income','beertax','baptist','mormon','drinkage','dry','youngdrivers','miles','gsp')]);


# Rewrite the functions in Chapter 2
linear.model<-function(beta,covariate){
  # beta: a 2-vector, where the first entry is the intercept
  beta=as.matrix(beta,nrow=length(beta))
  yout=covariate%*%beta[-1]+beta[1];
  return(yout);
  # Note: this function only works with simple linear regression model
  # How would you generalize it?
}
sum.of.squares<-function(beta,covariate,outcome){
  yout=linear.model(beta=beta,covariate=covariate);
  res=outcome-yout;
  sos= sum(res^2);
  return(sos)
}
fit.linear.model<-function(covariate,outcome){
  X=cbind(1,covariate);
  
  beta.fit=solve( t(X)%*%X )%*%t(X)%*%outcome;
  return(beta.fit)
}

beta.hat=fit.linear.model(covariate=X,outcome=y)

# Compare with lm()
fit.lm=lm(y~X+1);
beta.hat
```

```{r}
# Hat matrix 
calculate.hat<-function(covariate){
  X=cbind(1,covariate);
  P=X%*%solve(t(X)%*%X)%*%t(X);
  return(P)
}
P=calculate.hat(covariate=X);

# Verify the properties of the hat matrix:
## Residuals 
resid= y-linear.model(beta=beta.hat,covariate=X);
resid.hat = (diag(length(y))-P)%*%y;
sum((resid-resid.hat)^2)

## Idempotent 
max( abs(P%*%P-P))

## Orthogonality 
max(abs((diag(length(y))-P)%*%X))
```

```{r}
# Sum of squares with the hat matrix:
n=length(y);J=matrix(1,n,n);I=diag(n)
total.sum.of.squares= t(y)%*%(I-J/n)%*%y
explained.sum.of.squares=t(y)%*%(P-J/n)%*%y
residual.sum.of.squares=t(y)%*%(I-P)%*%y

# Check if it is true
sum.of.squares(beta.hat,covariate = X,outcome = y)
```

```{r}
# R^2
R.sq = explained.sum.of.squares/total.sum.of.squares;

# Through more variables into X
# These variables are meaningless
set.seed(1)
Z=matrix(rnorm(5*n),nrow=n,ncol=5);
X.Z=cbind(X,Z);

P.Z=calculate.hat(covariate=X.Z);

R.sq.Z= t(y)%*%(P.Z-J/n)%*%y/t(y)%*%(I-J/n)%*%y;

R.sq.Z
R.sq

# Hence the adjusted R^2:
R.sq.adj=1-(residual.sum.of.squares/(n-dim(X)[2]-1) )/(total.sum.of.squares/(n-1));

R.sq.Z.adj=1-(t(y)%*%(I-P.Z)%*%y/(n-dim(X.Z)[2]-1) )/(t(y)%*%(I-J/n)%*%y/(n-1));

```


## Underfitting and overfitting 


<b>Reading materials</b>:  Slides 142 - 152 in STA108_LinearRegression_S20.pdf. 


```{r}
## Underfitting
simulate.one.instance<-function(beta.true,X,Z){
  eta=c(3,4);
 n= dim(X)[1];
  noise = rnorm(n)/2;
  y.XZ=X%*%beta.true+Z%*%eta + noise;
  
  beta.full = fit.linear.model(covariate=cbind(X,Z),outcome=y.XZ)
  beta.under= fit.linear.model(covariate=X,outcome=y.XZ)
  
  return(cbind(beta.full[2:3]-beta.true,beta.under[2:3]-beta.true))
}

beta.true=c(1,2)
n=100;
X=matrix(rnorm(2*n),ncol=2)
Z=matrix(rnorm(2*n),ncol=2)
  
set.seed(1);
N.sim=10000;
under.sim=replicate(N.sim,simulate.one.instance(beta.true,X,Z))
# Fitting the true model gives

apply(under.sim[1,,],1,mean) # Non-zero bias

# Does it contradict the claim about precision variable?
# Try another simulation:
simulate.one.instance.pre<-function(beta.true){
  eta=c(3,4);
n=100;
X=matrix(rnorm(2*n),ncol=2)
Z=matrix(rnorm(2*n),ncol=2)
  noise = rnorm(n)/2;
  y.XZ=X%*%beta.true+Z%*%eta + noise;
  
  beta.full = fit.linear.model(covariate=cbind(X,Z),outcome=y.XZ)
  beta.under= fit.linear.model(covariate=X,outcome=y.XZ)
  
  return(cbind(beta.full[2:3]-beta.true,beta.under[2:3]-beta.true))
}

beta.true=c(1,2)
  
set.seed(1);
N.sim=10000;
pre.sim=replicate(N.sim,simulate.one.instance.pre(beta.true))

apply(pre.sim[1,,],1,mean) # almost zero bias
```

```{r}
### Overfitting
# Just modify the code slightly..
simulate.one.instance<-function(beta.true,X,Z){
  eta=c(3,4);
 n= dim(X)[1];
  noise = rnorm(n)/2;
  y.X=X%*%beta.true+ noise;
  
  beta.over= fit.linear.model(covariate=cbind(X,Z),outcome=y.X)
  beta.full= fit.linear.model(covariate=X,outcome=y.X)
  
  return(cbind(beta.full[2:3]-beta.true,beta.over[2:3]-beta.true))
}

beta.true=c(1,2)
n=100;
X=matrix(rnorm(2*n),ncol=2)
Z=matrix(rnorm(2*n),ncol=2)
  
set.seed(1);
N.sim=10000;
over.sim=replicate(N.sim,simulate.one.instance(beta.true,X,Z))
# Fitting the true model gives

apply(over.sim[1,,],1,mean) # No bias


```

## Sampling distribution and inference 


<b>Reading materials</b>:  Slides 153 - 165 in STA108_LinearRegression_S20.pdf. 

We will skip the part for multivariate normal distribution. You will learn more about this in STA 135: Multivariate Data Analysis.  

### ANOVA

```{r}
attach(dat.advertising)
full.model=lm(sales~TV+radio+newspaper);
reduced.model=lm(sales~TV);
anova(reduced.model,full.model)

RSS.Reduced<-sum((summary(reduced.model)$residuals)^2);
RSS.Full<-sum((summary(full.model)$residuals)^2);
df.Reduced <- summary(reduced.model)$df[2];
df.Full <- summary(full.model)$df[2];

tF<-((RSS.Reduced-RSS.Full)/(df.Reduced-df.Full))/(RSS.Full/df.Full);
tF
```


### Wald test


Suppose that we want to test the null hypothesis
$$H_0: \beta_1-2\beta_2 = 0 \ {\rm v.s.}\ H_a:  \beta_1-2\beta_2 \neq 0,$$
  where $\beta_1$ is the regression coefficient for `TV` and $\beta_2$ is the regression coefficient for `radio`.


```{r}
full.model=lm(sales~TV+radio+newspaper);
covariance.LSE=vcov(full.model); # Calculate the covariance for the least squares estimators
coef.LSE=coef(full.model);
covariance.test = covariance.LSE[2:3,2:3]; # Extract the submatrix of covariance for beta1 and beta2
coef.test=coef.LSE[2:3];
R=matrix(0,1,2);
R[1,1]=1;R[1,2]=-2;

test.stat=t(R%*%coef.test)%*%solve(R%*%covariance.test%*%t(R))%*%(R%*%coef.test);
p.value=1-pchisq(test.stat,df=1);
```


We can verify that the Wald test yields similar results with the F-test using `anova`.
$$H_0: \beta_1=\beta_2 = 0 \ {\rm v.s.}\ H_a:  \beta_1\neq 0, \beta_2 \neq 0.$$

```{r}

  
full.model=lm(sales~TV+radio+newspaper);
covariance.LSE=vcov(full.model); # Calculate the covariance for the least squares estimators
coef.LSE=coef(full.model);
covariance.test = covariance.LSE[2:3,2:3]; # Extract the submatrix of covariance for beta1 and beta2
coef.test=coef.LSE[2:3];
R=matrix(0,2,2);
R[1,1]=1;R[2,2]=1;

test.stat=t(R%*%coef.test)%*%solve(R%*%covariance.test%*%t(R))%*%(R%*%coef.test);
p_value=1-pchisq(test.stat,df=2);


reduced_model=lm(sales~newspaper);
anova(reduced.model,full.model)

```

