---
title: "Lecture script"
date: "April 13th 2020"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(include=TRUE,echo = TRUE, eval=FALSE, message=FALSE,warning=FALSE)
```

# Simple linear regression model 

\[
z_i = x_i \beta_1 +\beta_0 +\epsilon_i, \quad  i=1,\ldots, n
\]

Observed data:

- $(x_i, z_i)$ is the $i$th sample or observation among the $n$ samples ($i=1,2,..., n$)
- $z_i$ (which is $y_i$ in the slides) is the reponse, or dependent variable of the $i$th observation/sample 
- $x_i$ is the covariate, explanatory variable, or indepedent variable of the $i$th sample  
- $n$ is the sample size (number of observations)

Unknown (but fixed) parameters:

- $\beta_1$ is the (regression) slope, or regression coefficient of $x$ 
- $\beta_0$ is the intercept (on y-axis), or  (not very commonly used) regression coefficient of the constant 

Unknown random variable:
- $\epsilon_i$ is the noise or error term of the $i$th sample 

Assumptions on $\{\epsilon_i\}_{i=1}^{n} \equiv \{\epsilon_1,\epsilon_2,\ldots, \epsilon_n\}$

- $\{\epsilon_i\}_{i=1}^{n}$ are independently, and identically distributed (i.i.d.)
- $\mathbb{E}[\epsilon_i]=0$ 
- ${\rm var}(\epsilon_i)=\sigma^2$ 

- $\{\epsilon_i\}_{i=1}^{n}$ are independent with $\{x_i\}_{i=1}^n$ (not considered in this class, because the independent variables are assumed to be constant)


# Advertising data 

```{r}
dat.advertising=read.csv('../data/advertising.csv');

# Visualize the data set:
library(ggplot2)
library(GGally)

ggpairs(dat.advertising)
```


\[
z_i = x_i \beta_1 +\beta_0 +\epsilon_i, \quad  i=1,\ldots, n
\]

Observed data:

- $(x_i, z_i)$ is the $i$th sample or observation among the $n$ samples ($i=1,2,..., n$)
- $z_i$ (which is $y_i$ in the slides) is the reponse, which is the `sales` in this dataset 
- $x_i$ is the covariate, which is the `TV` (advertising budget) in this data set
- $n=200$ is the sample size (number of observations)

```{r}
# Fit a simple linear regression
# with sales as the response, and TV as the independent variable (covariate)
fit.advertising =lm(sales~TV+1,data=dat.advertising)
summary(fit.advertising)
```

We will denote the estimators as $\hat{\beta}_1$ and $\hat{\beta}_0$.

```{r}
# Draw the fitted line using ggplot2
ggplot(data = dat.advertising) + 
  geom_point(mapping = aes(x = TV, y = sales)) +
  geom_line(data = fortify(fit.advertising ), aes(x = TV, y = .fitted),color='red')
```

```{r}
ggplot(data = dat.advertising,aes(x = TV, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

# Flu shot 

```{r}

dat.flu = read.table("../data/flu240.txt", header = TRUE)

# Take a look at size of the data before proceed:
dim(dat.flu)
colnames(dat.flu)
# Basic visulization:
ggpairs(dat.flu) # This might take a while..
```

```{r}
# Jittering
ggplot(dat.flu, aes(x=treatment.assigned, y=outcome)) + 
  geom_point()

ggplot(dat.flu, aes(x=treatment.assigned, y=outcome)) + 
  geom_jitter(width = 0.25)
```

```{r}
ggplot(dat.flu, aes(x=treatment.assigned, y=outcome)) + 
  geom_jitter(width = 0.25)+
  geom_smooth(method = "lm")
```

# Project STAR
